Canopy leaf area of a mature evergreen *Eucalyptus* woodland does not respond to elevated atmospheric [CO~2~] but tracks water availability

### 2015-08-19 manuscript.Rmd
Many minor fixes

\
*Corresponding author*: 
Remko Duursma
E: remkoduursma@gmail.com
T: +61(0)422096908

@@

\
In ecosystems where water availability limits productivity, we expect that LAI responds especially strong to eC~a~ [@Donohue2009]. This is because eC~a~ leads to reduced stomatal conductance, and hence water use per unit leaf area [@Warren2010] leading to increased soil water availability. These 'water savings' can be expected to lead to an increase in LAI in water-limited ecosystems, consistent with a tight coupling of LAI with precipitation across gradients [@Ma2013; @Jin2013; @Ellis2008]. In support of this prediction, Donohue et al. [-@Donohue2013] attributed the observed global increase in LAI in arid and semi-arid areas over the last 30 years to rising atmospheric [CO~2~]. Based on the same water-savings principle, we predicted that LAI should increase under eC~a~ in water-limited forests.

\
The responses of LAI to fluctuations in water availability are qualitatively well-known [@Pook1985; @Specht1989; @Whitehead2004; @Pook1997; @Limousin2009], but poorly quantified. The strong relationship between LAI and rainfall observed across Australia [@Ellis2008; @Specht1989; @Broich2014], and other continents [@Donohue2013; @Grier1977; @Jin2013] represents the adaptation of vegetation structure and density to long-term climate conditions. However, local LAI is also highly responsive to water availability at shorter time scales [@Pook1985; @Pook1997], driven by the dynamics of leaf turnover [@Misson2011] and leaf flushing and expansion [@Muller2011], which are both highly responsive to soil water availability. Understanding dynamics of LAI in response to water availability and its potential interactions with eC~a~ is crucial if we are to improve phenology models in water-limited ecosystems [@Misson2011].

\
We estimated LAI from near-continuous measurements of canopy transmittance in a *Eucalyptus*-dominated mature woodland exposed to eC~a~, the EucFACE experiment. The strong seasonal and interannual variability in precipitation at the site allowed us to quantify the dynamics of LAI and its components (litter fall and production) in response to water availability. We hypothesize that eC~a~ would result in an increase in LAI, particularly during periods of water limitation. In addition, we expect that the low LAI that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allow a positive response of LAI to eC~a~ either via an increase in the rate of leaf production or leaf retention (leaf lifespan). Our second objective is to quantify the temporal dynamics of LAI, in particular in relation to fluctuating water availability.

@@

## Site description and experimental design
The EucFACE experiment is located in a remnant patch  of native Cumberland Plain woodland [@Tozer2003] on an ancient alluvial floodplain in western Sydney (Australia, 33^o^ 37'S, 150^o^ 44'E, 30 m a.s.l.). The vegetation consists of a native woodland undisturbed (unmanaged) for ca. 70 years and maintained in this location since the 1880's. The site is characterized by a humid temperate-subtropical transitional climate with a mean annual temperature of 17 ^o^C and mean annual precipitation (P) of 800 mm (Bureau of Meteorology, station 067105 in Richmond, NSW Australia; http://www.bom.gov.au). An estimate of potential evapotranspiration (PET) at the site (1350 mm) was from the Global-PET database [http://www.cgiar-csi.org, @Zomer2008], giving a ratio of P/PET of ca. 0.59. In addition, over the last 30 years, monthly PET has exceeded monthly P 78% of the time, indicating very frequent water limitation.

\
At the site, the soil at the surface is a loamy sand with > 75% sand content, and from ca. 50 to 300 cm depth, soils are a sandy clay loam with > 30% silt and clay. The soil is nutrient poor, and a strong limitation of P on tree growth at the site has recently been demonstrated [@Crous2015]. Groundwater depth is 12.6 – 12.9 m below the soil surface. The vegetation consists of an open woodland (600-1000 trees ha^-1^, basal area before treatment, BA = `r f(mean(ba$BA))`, SE = `r f(sd(ba$BA)/sqrt(6))` m^2^ha^-1^, n = 6 plots), with a single dominant species (*Eucalyptus tereticornis* Sm.) and some scattered individuals of *E. amplifolia* Naudin.  The understory stratum is  dominated by the grass *Microlaena stipoides* (Labill.) R.Br., with lower densities of forbs of *Commelina cyanea* R.Br., *Lobelia purpurascens* R.Br., *Dichondra repens* J.R.Forst. & G.Forst and occasional shrubs of *Breynia oblongifolia* (Mull.Arg.) Mull.Arg. and *Bursaria spinosa* Cav.

\
Within the EucFACE site, there are six 25-m diameter circular plots (designated hereafter as 'rings'). Each ring comprises 32 vent pipes in a cylindrical frame extending above the tree canopy to 28 m high. Each vertical pipe is subtended by a control valve and is connected to a circular plenum [@Hendrey1999] through which a computer-controlled system modulates release of diluted CO~2~ into the vegetation according to a proportional-integral-differential type of algorithm [@Lewin2009]. The vegetation within three randomly selected rings has been exposed to a CO~2~ mole fraction higher than ambient since September 2012, whereas the other three rings were identical to the CO~2~ treatment rings in infrastructure and instrumentation but received ambient C~a~, and are used as control plots. The target diurnal C~a~ for the rings exposed to eC~a~ was 150 $\mu$mol CO~2~ mol^-1^ above ambient C~a~ in full operation mode. However, C~a~ was increased gradually for six months at a rate of ca. 30 $\mu$mol CO~2~ mol^-1^ month^-1^ until full operation mode and the target eC~a~ was reached in February 2013 (Fig. 1a).

@@

## Terminology
In the following, we use the term 'leaf' area index (LAI) although the canopy and FACE rings contain non-leaf shading elements, i.e. woody stems and infrastructure. This avoids the need to switch between different terminologies when comparing temporal changes in LAI to litter production, and absolute values between different methods. The temporal changes in LAI are due to leaf area dynamics only, because infrastructure area is constant, and woody stem area is very nearly constant over the two-year study period since no tree mortality has occurred.

@@

## LAI from diffuse transmittance
In each of the six FACE rings, three photosynthetically active radiation (PAR) sensors (LI-190, Li-cor, Lincoln, NE, U.S.A.) were installed on metal posts at 1 m height. Data were logged on data loggers (CR3000, Campbell Scientific Australia, Townsville, Australia) at minutely intervals from 15 October 2012, and automatically uploaded every day to a data management system. At each ring, another PAR sensor was installed above the canopy, on a mast extending above a scaffolding tower in the centre of each ring at 23.5 m height. All PAR data were averaged in 30-min time steps. At a nearby flux tower site (within 2 km), we measured the diffuse fraction (F~diff~) of PAR at 30-min time step using a BF5 Sunshine sensor (Delta-T Instruments, Cambridge, UK) installed on a tower extending ca. 5 m above the canopy. For determination of LAI, we used only highly diffuse conditions, where F~diff~ > 0.98. We further trimmed the data by excluding low solar e levation (< 10^o^), because measured canopy transmittance was more variable at these low solar angles. Finally canopy transmittance was calculated for every 30-min time step as,

@@

\
We then estimated leaf area index (LAI) from $\tau_d$ ($LAI_\tau$), following Campbell and Norman [-@Campbell2000]. We assumed that leaves had a spherical leaf angle distribution, and leaves were randomly distributed in space. We recognize that these assumptions are very likely violated, but we account for this bias by calibrating the LAI estimate to leaf litter production as further detailed below. Diffuse transmittance of a canopy is a function of the LAI by integrating the Beer-Lambert law across the entire sky hemisphere, and accounting for differences in extinction coefficient with different viewing angles (because of leaf angle and longer path length through the canopy at low solar angles).

@@

where $\theta$ is the solar zenith angle, G($\theta$) is the foliage projection function in the direction of the beam [@Campbell2000]. Because we assumed a spherical leaf angle distribution, G($\theta$) = 0.5 [@Stenberg2006]. We solved numerically for LAI in Eq. 2, using a smooth integration over $\theta$, yielding $LAI_\tau$.

\
To overcome violations in the assumptions underlying Eq. 2, we calibrated $LAI_\tau$ against cumulative litter production during a 4-month drought period in 2013 (indicated in Fig. 1). During this period (14 July 2013 to 12 Nov 2013), no leaf growth occurred but LAI decreased steadily (and thus $\tau_d$ increased steadily, see Fig. 1). During the drought, only 42 mm precipitation was recorded, which was the second driest July-October period recorded at a nearby (<2km) weather station since 1881 (compared to a long-term average of 188 mm). A total of n = 11 days with sufficiently diffuse conditions to estimate $\tau_d$  were available during this period. Cumulative LAI leaf litter production ($P_L$, expressed in units of LAI) should match the observed decrease in LAI estimated from canopy transmittance. We calibrated Eq. 3,

@@

where *c* is a calibration parameter, $L_{\tau,t=0}$ is $LAI_\tau$ at the start of the drought period. We used linear regression without intercept to estimate *c* in Eq. 4 as `r f(calib)` (SE = `r f(tidy(caliblm)$std.error)`). It should be stressed that even after calibration with litter production, $LAI_\tau$ still includes stem and infrastructure area (see Terminology). However, this approximately constant area cancels out in the left-hand side of Eq. 3. We used a single calibration constant for all rings, because we did not find significant differences between the rings in terms of either litter production or change in $\tau_d$ during the 2013 drought period (see Fig. S1).

@@

\
To test for effects of C~a~ treatment on LAI and litter production, we used linear-mixed effects models with FACE ring as the random effect, and C~a~ treatment and date as factor variables. The random effect accounts for the repeated measures nature of the data. For LAI, we further used pre-treatment basal area (BA) of each FACE ring as a covariate. Because the EucFACE woodland is well established, the differences in BA are not due to C~a~ treatment effects, but instead represent natural variation within the woodland.

**More details here;package, which anova for Table 1; and now we use BA as a covariate  **

\ To compare the dynamics of LAI with litter production, we calculated the change in LAI for every time period when litter was collected. We did this by fitting a generalized additive model to LAI for each ring separately (similar to that shown for $\tau_d$ in Fig. 1), and calculating the change in LAI for each period ($\Delta L_\tau$). Next, we estimated gross leaf production from measured litter production (P~L~) and temporal dynamics of $LAI_\tau$. For each litter collection period, the net change in $LAI_\tau$ was estimated using fitted generalized additive models to allow interpolation from a smoothed timeseries. Gross leaf production was then calculated as the sum of the net change in $LAI_\tau$ and the amount of leaf litter produced during that period. We estimated mean leaf lifespan (LL) as the ratio of $LAI_\tau$ to annual total P~L~.

\
We quantified the relationship between timing of leaf flushing and environmental drivers. During the `r nummon` month study period, we observed four leaf flushing events in the field, with one before estimates of $LAI_\tau$ were available. These events occurred in spring and summer, with no clear seasonal cycle (October 2012, January 2013, December 2013 and January 2015). These leaf growth events were compared to the dynamics of LAI as estimated from a generalized additive model fitted to the average $LAI_\tau$ across the six FACE rings. From the fitted model, we estimated the numerical derivative of LAI (dL/dt), as well as its confidence interval following Simpson [-@Simpson2014; see also @Curtis2014].

\
The entire analysis presented in this paper will be made available as a *git* repository (http://www.github.com/remkoduursma/XXXXX), published on figshare (doi:XXXX). The repository includes the raw data needed to run the analyses, the source code for the analysis and manuscript (which was written in markdown, with the rmarkdown package in Rstudio, which uses Pandoc for conversion).

# Results
Canopy transmittance ($\tau_d$) varied between `r f(min(df$Gapfraction.mean))` and `r f(max(df$Gapfraction.mean))`, with much temporal variation, and substantial absolute differences between the individual FACE rings (Fig. 1).  However, temporal trends were very similar between the rings. Since the understorey comprises mostly grasses and forbs, and the PAR sensors were installed at 1 m height, this variation is due to the overstorey trees only.

\
An outbreak of psyllids (*Cardiaspina* sp.), a tiny sap-feeding insect, resulted in rapid loss of canopy leaf area in one of the rings (ring 6), starting in September 2014 [for description of a similar outbreak in *Eucalyptus moluccana*, see @Hall2015; pers. comm. Andrew Gherlenda, Markus Riegler]. This rapid decrease in LAI was evident in a quick increase in $\tau_d$ (Fig. 1). However, it recovered rapidly after rainfall in December 2014, resulting in a lower $\tau_d$ than before the insect attack.

\
We found no differences in LAI between the C~a~ treatments, as shown by consistent overlap of the 95% confidence intervals of the generalized additive model (thus, p > 0.05 at any point in time, Fig. 2). A linear mixed-effects model with C~a~, time and pre-treatment basal area (Fig. 3) as fixed effects, and FACE ring as a random effect also showed no significant effect of C~a~ on LAI (p **fix this**).

\
Litter production correlated significantly with $\Delta L_\tau$ when the latter was separated into periods where $\Delta L_\tau$ > 0 (net leaf growth) and $\Delta L_\tau$ <  0 (net leaf loss) (Fig. 3). During periods of net leaf loss, we found a significant correlation (with a slope of `r f(tidy(lmlitdown)$estimate[2])`, SE = `r f(tidy(lmlitdown)$std.error[2])`, n = `r glance(lmlitdown)$df.residual+2`), close to the 1:1 line, as expected since no leaf growth was observed in the field during any period where LAI was decreasing (see also Fig. 6). During periods of net leaf growth, we found a significant correlation (with a slope of `r f(tidy(lmlitup)$estimate[2])`, SE = `r f(tidy(lmlitup)$std.error[2])`, n = `r glance(lmlitup)$df.residual+2`) between $\Delta L_\tau$ and litter production (Fig. 4).

\
The temporal dynamics of litter and leaf production are shown in Fig. 4, and demonstrate the irregular temporal pattern in leaf production and P~L~. We found no significant difference in either litter or leaf production between C~a~ treatments at any time point (p > 0.2 for both, assessed with linear-mixed effects models).

\
Next, we related P~L~ to average $LAI_\tau$ (during the litter collection period) across the six FACE rings (Fig. 5a). If leaf lifespan (LL) is constant, P~L~ would be linearly related to $LAI_\tau$. Instead, we found that in eC~a~, P~L~ was lower at a given $LAI_\tau$ as compared to aC~a~ (Fig. 6a). We found a significant (p `r pval(LLtest$p.value)`) increase in LL with elevated C~a~ (Fig. 5b), with LL = `r f(LLtest$estimate1)`yr in aC~a~ and `r f(LLtest$estimate2)`yr in eC~a~.

\
The dynamics of $LAI_\tau$ from the sensor record matched leaf growth events observed in the field quite well (Fig. 6a). All four leaf flushing events corresponded to significant positive dL/dt. One brief significant positive dL/dt (June 2013) did not correspond to an observed leaf flushing event, but it was small in magnitude. It is also obvious that every leaf flush event occurred shortly after a significant rainfall event (Fig. 3 and 5), when the soil water content in the top 80 cm increased to ca. > 0.2 m^3^ m^-3^. Two periods of sustained decrease in LAI (red lines in Fig. 6b, during winter of 2013 and 2014) corresponded to periods of low soil water availability.

@@

## Response of LAI to elevated [CO~2~]
We report on `r nummon` months of continuous leaf area index (LAI) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated CO~2~ (eC~a~), including 3 months of ramp-up in C~a~ from November 2012 to January 2013. We predicted that LAI would increase in response to eC~a~ as a result of enhanced CO~2~ assimilation for the entire canopy and increased soil water availability. We also expected that this eucalypt woodland, with its inherently low LAI [@Whitehead2004], would be even more responsive to eC~a~ than higher-LAI and more closed-canopy stands [@Norby2011]. Counter to our expectations but similar to Norby et al. [-@Norby2003], we found no significant effect of elevated C~a~ on LAI. Because estimated LL averaged `r round(mean(ba$LL),1)` yrs (Fig. 5), the canopy has regrown at least twice during the study period, giving ample opportunity for eC~a~ effects to occur. It could be argued that the change in LAI in response to eC~a~ was too small to be detected. Natural variation in LAI among plots in native woodland is indeed a challenge to manipulative studies with low replication, such as FACE experiments. To increase sensitivity and statistical robustness, we used a novel method to account for pre-treatment differences by using stand basal area as a covariate. Using this method, we estimated that we can detect a difference of only a `r format(Ldetect,digits=1)`% in LAI. Based on this small detectable difference, we conclude that the lack of a significant C~a~ effect is not due to a lack of power, but a true lack of response to elevated C~a~.

@@

\
This lack of a response to elevated C~a~ counters our expectations and those in previous publications [@Norby2003; @McCarthy2007] for a number of reasons. First, in a review of forest FACE sites, Norby and Zak [-@Norby2011] showed that low LAI stands were more responsive to eC~a~ than stands with high LAI. They argued that high LAI canopies would not benefit from further increases in LAI, because it would lead to only very small increases in total absorbed radiation. The mean LAI at the EucFACE was ca. `r f(mean(ba$LAI.mean))` m^2^m^-2^, which is low compared to a global compendium of forest LAI [@Iio2013], and the lowest LAI of any forest FACE experiment [@Norby2011], but in line with a range of values for native *Eucalyptus* forests and woodlands reported by Whitehead and Beadle [-@Whitehead2004] and Iio et al. [-@Iio2013]. Based on this LAI estimate and the relationship across FACE sites reported by Norby and Zak [-@Norby2011], we expect about a 30% increase in LAI under eC~a~. In contrast to other forest FACE sites, the EucFACE is strongly limited by nutrient availability, in particular phosphorus [@Crous2015; @Nielsen2015]. Hence, it may be too early to draw general conclusions across FACE sites in terms of LAI response.

>>>>

\
*Corresponding author*: 
Remko Duursma
\
E: remkoduursma@gmail.com
\
T: +61(0)422096908

@@

\
In ecosystems where water availability limits productivity, we expect an especially large response of LAI to eC~a~ [@Donohue2009]. This is because eC~a~ leads to reduced stomatal conductance, and hence water use per unit leaf area [@Warren2010] leading to increased soil water availability. These 'water savings' can be expected to lead to an increase in LAI in water-limited ecosystems, consistent with a tight coupling of LAI with precipitation across gradients [@Ma2013; @Jin2013; @Ellis2008]. In support of this prediction, Donohue et al. [-@Donohue2013] attributed the observed global increase in LAI in arid and semi-arid areas over the last 30 years to rising atmospheric [CO~2~]. Based on the same water-savings principle, we predicted that LAI should increase under eC~a~ in water-limited forests.

\
The responses of LAI to fluctuations in water availability are qualitatively well-known [@Pook1985; @Specht1989; @Whitehead2004; @Pook1997; @Limousin2009], but poorly quantified. The strong relationship between LAI and rainfall observed across Australia [@Ellis2008; @Specht1989; @Broich2014], and other continents [@Donohue2013; @Grier1977; @Jin2013] represents the adaptation of vegetation structure and density to long-term climate conditions. However, local LAI is also highly responsive to water availability at shorter time scales [@Pook1985; @Pook1997], driven by the dynamics of leaf turnover [@Misson2011] and leaf flushing and expansion [@Muller2011]. Understanding dynamics of LAI in response to water availability and its potential interactions with eC~a~ is crucial if we are to improve phenology models in water-limited ecosystems [@Misson2011].

\
We estimated LAI from near-continuous measurements of canopy transmittance in a *Eucalyptus*-dominated mature woodland exposed to eC~a~, the EucFACE experiment. The strong seasonal and interannual variability in precipitation at the site allowed us to quantify the dynamics of LAI and its components (litter fall and production) in response to water availability. We hypothesize that eC~a~ would result in an increase in LAI, particularly during periods of water limitation. In addition, we expect that the low LAI that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allow a positive response of LAI to eC~a~ either via an increase in the rate of leaf production or leaf retention (leaf lifespan). Our second objective is to quantify the temporal dynamics of LAI and its components (litter fall and leaf production), in particular in relation to fluctuating water availability.

@@

## Site description and experimental design
The EucFACE experiment is located in a remnant patch  of native Cumberland Plain woodland [@Tozer2003] on an ancient alluvial floodplain in western Sydney (Australia, 33^o^ 37'S, 150^o^ 44'E, 30 m a.s.l.). The vegetation consists of a native woodland undisturbed (unmanaged) for ca. 70 years and maintained in this location since the 1880's. The site is characterized by a humid temperate-subtropical transitional climate with a mean annual temperature of 17 ^o^C and mean annual precipitation (P) of 800 mm (1881 - 2014, Bureau of Meteorology, station 067105 in Richmond, NSW Australia; http://www.bom.gov.au). An estimate of potential evapotranspiration (PET) at the site (1350 mm) was from the Global-PET database [1950 - 2000 climatology, http://www.cgiar-csi.org, @Zomer2008], giving a ratio of P/PET of ca. 0.59. In addition, over the last 30 years, monthly PET has exceeded monthly P 78% of the time, indicating frequent water limitation.

\
At the site, the soil at the surface is a loamy sand with > 75% sand content, and from ca. 50 to 300 cm depth, soils are a sandy clay loam with > 30% silt and clay. The soil is nutrient poor, and a strong limitation of P on tree growth at the site has recently been demonstrated [@Crous2015]. Groundwater depth is 12.6 - 12.9 m below the soil surface. The vegetation consists of an open woodland (600-1000 trees ha^-1^, basal area before treatment, BA = `r f(mean(ba$BA))`, SE = `r f(sd(ba$BA)/sqrt(6))` m^2^ha^-1^, n = 6 plots), with a single dominant species (*Eucalyptus tereticornis* Sm.) and some scattered individuals of *E. amplifolia* Naudin.  The understory stratum is  dominated by the grass *Microlaena stipoides* (Labill.) R.Br., with lower densities of forbs of *Commelina cyanea* R.Br., *Lobelia purpurascens* R.Br., *Dichondra repens* J.R.Forst. & G.Forst and occasional shrubs of *Breynia oblongifolia* (Mull.Arg.) Mull.Arg. and *Bursaria spinosa* Cav.

\
Within the EucFACE site, there are six 25-m diameter circular plots (designated hereafter as 'rings'). Each ring comprises 32 vent pipes in a cylindrical frame extending above the tree canopy to 28 m high. Each vertical pipe is subtended by a control valve and is connected to a circular plenum [@Hendrey1999] through which a computer-controlled system modulates release of diluted CO~2~ into the vegetation according to a proportional-integral-differential type of algorithm [@Lewin2009]. The vegetation within three randomly selected rings has been exposed to a CO~2~ mole fraction higher than ambient since September 2012, whereas the other three rings were identical to the CO~2~ treatment rings in infrastructure and instrumentation but received aC~a~, and are used as control plots. The target diurnal C~a~ for the rings exposed to eC~a~ was 150 $\mu$mol CO~2~ mol^-1^ above aC~a~ in full operation mode. However, C~a~ was increased gradually for six months at a rate of ca. 30 $\mu$mol CO~2~ mol^-1^ month^-1^ until full operation mode and the target eC~a~ was reached in February 2013 (Fig. 1a).

@@

## Terminology
In the following, we use the term leaf area index (LAI) although the canopy and FACE rings contain non-leaf shading elements, i.e. woody stems and infrastructure. This avoids the need to switch between different terminologies when comparing temporal changes in LAI to litter production, and absolute values between different methods. The temporal changes in LAI are due to leaf area dynamics only, because infrastructure area is constant, and woody stem area is very nearly constant over the two-year study period since no tree mortality has occurred.

@@

## LAI from diffuse transmittance
In each of the six FACE rings, three photosynthetically active radiation (PAR) sensors (LI-190, Li-cor, Lincoln, NE, U.S.A.) were installed on metal posts at 1 m height. Data were logged on data loggers (CR3000, Campbell Scientific Australia, Townsville, Australia) at minutely intervals from 15 October 2012, and automatically uploaded every day to a data management system. At each ring, another PAR sensor was installed above the canopy, on a mast extending above a scaffolding tower in the centre of each ring at 23.5 m height. All PAR data were averaged at 30-min time steps. At a nearby flux tower site (within 2 km), we measured the diffuse fraction (F~diff~) of PAR at 30-min time step using a BF5 Sunshine sensor (Delta-T Instruments, Cambridge, UK) installed on a tower extending ca. 5 m above the canopy. For determination of LAI, we used only highly diffuse conditions, where F~diff~ > 0.98. We further trimmed the data by excluding low solar elevation (< 10^o^), because measured canopy transmittance was more variable at these low solar angles. Finally canopy transmittance was calculated for every 30-min time step as,

@@

\
We then estimated leaf area index (LAI) from $\tau_d$ following Campbell and Norman [-@Campbell2000]. We assumed that leaves had a spherical leaf angle distribution, and leaves were randomly distributed in space. We recognize that these assumptions are very likely violated, but we account for this bias by calibrating the LAI estimate to leaf litter production as further detailed below. Diffuse transmittance of a canopy is a function of the LAI by integrating the Beer-Lambert law across the entire sky hemisphere, and accounting for differences in extinction coefficient with different viewing angles (because of leaf angle and longer path length through the canopy at low solar angles).

@@

where L is LAI, $\theta$ is the solar zenith angle, G($\theta$) is the foliage projection function in the direction of the beam [@Campbell2000]. Because we assumed a spherical leaf angle distribution, G($\theta$) = 0.5 [@Stenberg2006]. We solved numerically for LAI in Eq. 2, using a smooth integration over $\theta$, yielding LAI.

\
To overcome violations in the assumptions underlying Eq. 2, we calibrated LAI against cumulative litter production during a 4-month drought period in 2013 (indicated in Fig. 1). During this period (14 July 2013 to 12 Nov 2013), no leaf growth occurred but LAI decreased steadily (and thus $\tau_d$ increased steadily, see Fig. 1). During the drought, only 42 mm precipitation was recorded, which was the second driest July-October period recorded at a nearby (<2km) weather station since 1881 (compared to a long-term average of 188 mm). A total of n = 11 days with sufficiently diffuse conditions to estimate $\tau_d$  were available during this period. Cumulative LAI leaf litter production ($P_L$, expressed in units of LAI) should match the observed decrease in LAI estimated from canopy transmittance. We calibrated Eq. 3,

@@

where *c* is a calibration parameter, $L_{\tau,t=0}$ is LAI at the start of the drought period. We used linear regression without intercept to estimate *c* in Eq. 4 as `r f(calib)` (SE = `r f(tidy(caliblm)$std.error)`). It should be stressed that even after calibration with litter production, LAI still includes stem and infrastructure area (see Terminology). However, this approximately constant area cancels out in the left-hand side of Eq. 3. We used a single calibration constant for all rings, because we did not find significant differences between the rings in terms of either litter production or change in $\tau_d$ during the 2013 drought period (see Fig. S1).

@@

\
To test for effects of C~a~ treatment on LAI and litter production, we used linear-mixed effects models with FACE ring as the random effect, and C~a~ treatment and date as factor variables. The random effect accounts for the repeated measures nature of the data. For LAI, we further used pre-treatment basal area (BA) of each FACE ring as a covariate (see Fig. S4). Because the EucFACE woodland is well established, the differences in BA are not due to C~a~ treatment effects, but instead represent natural variation within the woodland.

\
To compare the dynamics of LAI with litter production, we calculated the change in LAI for every time period when litter was collected. We did this by fitting a GAM to LAI for each ring separately (similar to that shown for $\tau_d$ in Fig. 1), and calculating the change in LAI for each litter fall collection period ($\Delta$LAI). Next, we estimated gross leaf production (G~L~) as the sum of the net change in LAI ($\Delta$LAI) and P~L~, because mass balance requires that $\Delta$LAI = G~L~ - P~L~. We estimated mean leaf lifespan (LL, yr) as the ratio of mean LAI (m^2^ m^-2^) for the entire litter fall collection period to total P~L~ (m^2^ m^-2^ yr^-1^).

\
We quantified the relationship between timing of leaf flushing and environmental drivers. During the `r nummon` month study period, we observed four leaf flushing events in the field, with one before estimates of LAI were available. These events occurred in spring and summer, with no clear seasonal cycle (October 2012, January 2013, December 2013 and January 2015). These leaf growth events were compared to the dynamics of LAI as estimated from a GAM fitted to the average daily LAI across the six FACE rings. From the fitted model, we estimated the numerical derivative of LAI (dL/dt), as well as its confidence interval following Simpson [-@Simpson2014, see also @Curtis2014].

\
The entire analysis presented in this paper will be made available as a *git* repository (http://www.github.com/remkoduursma/XXX), published on figshare (doi:XXX). The repository includes the raw data needed to run the analyses, the source code for the analysis and manuscript (which was written in markdown, using the 'rmarkdown' R package for conversion).

# Results
Canopy transmittance ($\tau_d$) varied between `r f(min(df$Gapfraction.mean))` and `r f(max(df$Gapfraction.mean))`, with much temporal variation, and substantial absolute differences between the individual FACE rings (Fig. 1).  However, temporal trends were very similar between the rings. Since the understorey comprises mostly grasses and forbs, and the PAR sensors were installed at 1 m height, this variation is due to the overstorey trees only. An outbreak of psyllids (*Cardiaspina* sp.), a tiny sap-feeding insect, resulted in rapid loss of canopy leaf area in one of the rings (ring 6), starting in September 2014 [for description of a similar outbreak in *Eucalyptus moluccana*, see @Hall2015; pers. comm. Andrew Gherlenda, Markus Riegler]. This rapid decrease in LAI was evident in a quick increase in $\tau_d$ (Fig. 1). However, it recovered rapidly after rainfall in December 2014, resulting in a lower $\tau_d$ than before the insect attack.

\
We found no differences in LAI between the C~a~ treatments, as shown by consistent overlap of the 95% confidence intervals of the generalized additive model (thus, p > 0.05 at any point in time, Fig. 2). A linear mixed-effects model with C~a~, time and pre-treatment basal area (Fig. 3) as fixed effects, and FACE ring as a random effect also showed no significant effect of C~a~ on LAI (p > 0.5).

\
Litter production (P~L~) correlated significantly with $\Delta$LAI when the latter was separated into periods where $\Delta$LAI > 0 (net leaf growth) and $\Delta$LAI <  0 (net leaf loss) (Fig. 3). During periods of net leaf loss, we found a significant correlation (with a slope of `r f(tidy(lmlitdown)$estimate[2])`, SE = `r f(tidy(lmlitdown)$std.error[2])`, n = `r glance(lmlitdown)$df.residual+2`), close to the 1:1 line, as expected since no leaf growth was observed in the field during any period where LAI was decreasing (see also Fig. 6). During periods of net leaf growth, we found a significant correlation (with a slope of `r f(tidy(lmlitup)$estimate[2])`, SE = `r f(tidy(lmlitup)$std.error[2])`, n = `r glance(lmlitup)$df.residual+2`) between $\Delta$LAI and litter production (Fig. 4).

\
The temporal dynamics of litter and leaf production are shown in Fig. 4, and demonstrate the irregular temporal pattern in both leaf production and P~L~. We found no significant difference in either litter or leaf production between C~a~ treatments at any time point (p > 0.2 for both, assessed with linear-mixed effects models).

\
Next, we related total P~L~ to average LAI (during the litter collection period) across the six FACE rings (Fig. 5a). If leaf lifespan (LL) is constant, P~L~ would be linearly related to LAI. Instead, we found that in eC~a~, P~L~ was lower at a given LAI as compared to aC~a~ (Fig. 6a). We found a significant (p `r pval(LLtest$p.value)`) increase in LL with eC~a~ (Fig. 5b), with LL = `r f(LLtest$estimate1)`yr in aC~a~ and `r f(LLtest$estimate2)`yr in eC~a~.

\
The dynamics of LAI from the sensor record matched leaf growth events observed in the field quite well (Fig. 6a). All four leaf flushing events corresponded to significant positive dL/dt. One brief significant positive dL/dt (June 2013) did not correspond to an observed leaf flushing event, but it was small in magnitude. It is also obvious that every leaf flush event occurred shortly after a significant rainfall event (Fig. 3 and 5), when the soil water content in the top 80 cm increased to ca. > 0.2 m^3^ m^-3^. Two periods of sustained decrease in LAI (red lines in Fig. 6b, during winter of 2013 and 2014) corresponded to periods of low soil water availability.

@@

## Response of LAI to elevated [CO~2~]
We report on `r nummon` months of continuous leaf area index (LAI) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated CO~2~ (eC~a~), including 3 months of ramp-up in C~a~ from November 2012 to January 2013. We predicted that LAI would increase in response to eC~a~ as a result of enhanced CO~2~ assimilation for the entire canopy and increased soil water availability. We also expected that this eucalypt woodland, with its inherently low LAI [@Whitehead2004], would be even more responsive to eC~a~ than higher-LAI and more closed-canopy stands [@Norby2011]. Counter to our expectations but similar to Norby et al. [-@Norby2003], we found no significant effect of eC~a~ on LAI. Because estimated LL averaged `r round(mean(ba$LL),1)` yrs (Fig. 5), the canopy has regrown at least twice during the study period, giving ample opportunity for eC~a~ effects to occur. It could be argued that the change in LAI in response to eC~a~ was too small to be detected. Natural variation in LAI among plots in native woodland is indeed a challenge to manipulative studies with low replication, such as FACE experiments. To increase sensitivity and statistical robustness, we used pre-treatment stand basal area as a covariate. Using this method, we estimated that we can detect a difference of only a `r format(Ldetect,digits=1)`% in LAI. Based on this small detectable difference, we conclude that the lack of a significant C~a~ effect is not due to a lack of power, but a true lack of response to eC~a~.

@@

\
This lack of a response to eC~a~ counters our expectations and those in previous publications [@Norby2003; @McCarthy2007] for a number of reasons. First, in a review of forest FACE sites, Norby and Zak [-@Norby2011] showed that low LAI stands were more responsive to eC~a~ than stands with high LAI. They argued that high LAI canopies would not benefit from further increases in LAI, because it would lead to only very small increases in total absorbed radiation. The mean LAI at the EucFACE was ca. `r f(mean(ba$LAI.mean))` m^2^m^-2^, which is low compared to a global compendium of forest LAI [@Iio2013], and the lowest LAI of any forest FACE experiment [@Norby2011], but in line with a range of values for native *Eucalyptus* forests and woodlands reported by Whitehead and Beadle [-@Whitehead2004] and Iio et al. [-@Iio2013]. Based on this LAI estimate and the relationship across FACE sites reported by Norby and Zak [-@Norby2011], we expect about a 30% increase in LAI under eC~a~. In contrast to other forest FACE sites, the EucFACE is strongly limited by nutrient availability, in particular phosphorus [@Crous2015; @Nielsen2015]. Hence, it may be too early to draw general conclusions across FACE sites in terms of LAI response.

\
Stimulation of net leaf photosynthesis by eC~a~ (a common response, also observed at the EucFACE, Gimeno et al. *in press*) may not lead to increased net leaf growth, due to other constraints such as water or nutrient limitations. Net leaf growth is the net sum of allocation of photosynthates to leaf growth, and turnover of the leaf pool. Either process may itself be affected by eC~a~ [@DeKauwe2014]. We found a significant effect of eC~a~ on LL, with leaves living approximately `r round(abs(LLtest$estimate*12),1)` months longer in eC~a~ (Fig. 5) (an increase of `r round(100*with(LLtest,estimate2/estimate1 - 1),0)`%). Because we found no difference in LAI with C~a~ treatment, this suggests a compensating difference in the fractional allocation of net primary production to leaf growth. To further understand the response of biomass allocation to eC~a~ at the EucFACE, all components of the carbon balance must be carefully quantified, a task underway but not yet complete.
